Executive Constraints and Civil Conflict Onset

### 2020-05-21 paper.Rmd
Update stan file path in manuscript
Draft edits

Finally, figure \ref{fig:peace} plots the peace year intercepts.[^gp]
The strongest effect is found in the years immediately following an
occurrence of intrastate political violence resulting in an increased
risk for an additional onset.  Approximately a decade after the last
incidence of violence, the effect is drawn towards zero. The
additional spike around approximately 30 years follows an uptick in
conflict onsets in the dataset mostly among authoritarian regimes such
as, for example, the 1989 coup d'etat overthrowing the Stroessner
regime in Paraguay, the Saur Revolution in 1978 in Afghanistan, and
the start of the Nepalese Civil War in 1996. After 30 years, the
effect of each additional year of peace tends to be negative; however,
there is a very high degree of uncertainty in the estimates that
undermines any general conclusion regarding a pacifying effect.

@@

To interpret the results for the interaction term between the slopes
for executive constraints and GDP per Capita I calculate the marginal
effect of $\theta$, executive constraints, on the predicted
probability of civil conflict onset for different levels of GDP per
capita while holding constant the other covariates.  In this case, I
take the mean of GDP per capita growth, CINC, the proportion of
discriminated population, population density, mean elevation, and the
number of politically relevant ethnic groups. I also set the dummy
indicators for independence, major ongoing conflict, and neighbouring
conflict and the random intercepts for year and country to zero. For
peace years I take the intercept for the average number of peace years
rounded to the nearest integer, which turns out to be 19. The levels
of GDP per capita were selected based on the 10\%, 50\%, and 90\%
quantiles of the standardized log transformed values. These levels
roughly correspond to the GDP per capita of Rwanda in 2007 at
$`r gdp_rwa`$ 2011 US dollars, the Philippines in 2008 at $`r gdp_phl`$
2011 US dollars, and Portugal in 2009 at $`r gdp_po`$ 2011 US
dollars. In the dataset there are 60 conflict episode onsets below the
first level, 152 between the first and second, 55 between the second
and third, and 4 above the third.

```{r fig.cap = "\\label{fig:int} Interaction between executive constraints and GDP per capita with 80\\% CIs"}
# Interaction plot
theta_seq <- seq(-2.5, 2.5, .01)
state <- quantile(X[, "rgdpepc"], probs = c(0.1, .5, 0.9))

@@

From the results in figure \ref{fig:int} it is clear that the overall
effect of executive constraints when interacted with GDP per capita on
the predicted probability of conflict onset is negative. For countries
with a relatively high level of GDP per capita, going from the worst
to best in executive constraints decreases the predicted probability
of conflict onset from [1.5, 4.5] down to below 0.5. This effect
diminishes, however, for lower values of GDP per capita to the point
that executive constraints are practically indeterminate to the risk
of conflict onset for low-income country-years.

@@

Figure \ref{fig:phat} plots the densities of the first 50 posterior
draws for the predicted probabilities of conflict onset for two
groups, country-years that actually experienced an onset of a civil
conflict and country-years that did not. On average, the predicted
probabilities for the actual onset country-years are higher albeit
still extremely low. This can be seen in the summary statistics
reported in table \ref{tab:psum} for the posterior means of the
predicted probabilities. For onset country-years the interquartile
interval is only 0.0475 to 0.2 with a max of 0.516. To better compare
the predicted probabilities to the actual data, table
\ref{tab:confusion} presents a confusion matrix with a 0.2 cutoff,
which approximately corresponds to the third quartile in the summary
table.  `r signif((93 / (259 + 93)) * 100, 3)` percent of onsets and
`r signif((7902 / (7902 + 172)) * 100, 3)` percent of non-onsets are
correctly predicted. The false positive rate is
`r signif((172 / (7902 + 172)) * 100, 3)` percent and the false
negative rate is `r signif((259 / (93 + 259)) * 100, 3)` percent.

@@

```{r psum}
p_mean <- colMeans(p_hat)

df <- bind_rows(summary(p_mean[final.df$lepisode_onset == 1]),
                summary(p_mean[final.df$lepisode_onset == 0])) %>%
    mutate_all(~signif(., 3))

@@

```{r confusion}
m <- table(p_mean > .2, final.df$lepisode_onset)

@@

@@

```{r results = "hide"}
x <- evalmod(scores = p_mean, labels = final.df$lepisode_onset)

@@

Figure \ref{fig:roc} plots a receiver operating characteristic (ROC)
curve depicting the tradeoff between the true positive rate
(*sensitivity*) and the false positive rate --- $1 -
\mbox{specificity}$ where *specificity* refers to the true negative
rate --- for varying probability thresholds according to the posterior
means of the predicted probabilities. For example, for a threshold of
approximately 0.1, the model will correctly predict 50 percent of
onsets while misclassifying only 8 percent of non-onsets. The total
performance of the model can be summarized by the area under the curve
(AUC), which in this case is `r auc`.[^auc] However, for the present
analysis the results from the ROC curve are misleading since there is
a significant class imbalance in the dataset. Onsets comprise only 4
percent of all observations meaning that the false positive rate is
less sensitive to changing thresholds. Figure \ref{fig:roc} therefore
also plots a precision-recall (PRC) curve, which depicts the tradeoff
between *precision* --- the number of true positives divided by the
sum of the true positives and false positives --- and *recall* --- the
number of true positives divided by the sum of the true positives and
false negatives --- for varying thresholds. The advantage is that
*precision* is invariant to the number of non-onset years in the
dataset. From this plot it is clear that the precision of the model,
the ability to discriminate between onset country-years and false
predicted non-set country-years, drops drastically as the *recall* of
the model increases. The *AUC-PR* for this plot is only `r auc_pr`.
[^auc_pr]

@@

Taken together, the evidence suggests a fairly poor fit to the
underlying data. The confusion matrix demonstrates that even with a
relatively low cutoff, the model is still only able to accurately
predict 25 percent of conflict onsets. According to the ROC curve,
decreasing this threshold improves the true positive rate with a
relatively decent tradeoff in terms of the false positive rate, but as
stated, this is largely due to the imbalance in the outcome
variable. In terms of *precision*, the performance of the model
quickly degrades as the threshold decreases.

@@

```{stan, code = readLines("stan/model.stan"), eval = F, output.var = "x", echo = T}
```

>>>>

Finally, figure \ref{fig:peace} plots the peace year intercepts.[^gp]
The strongest effect is found in the years immediately following an
occurrence of intrastate political violence resulting in an increased
risk for an additional onset.  Approximately a decade after the last
incidence of violence, the effect is drawn towards zero. The
additional spike around approximately 30 years follows an uptick in
conflict onsets in the dataset mostly among authoritarian regimes such
as, for example, the 1989 coup d'etat overthrowing the Stroessner
regime in Paraguay, the Saur Revolution in 1978 in Afghanistan, and
the start of the Nepalese Civil War in 1996. After 30 years, the
effect of each additional year of peace tends to be negative; however,
there is a very high degree of uncertainty in the estimates that
undermines any general conclusion regarding a pacifying effect in the
model.

@@

To interpret the interaction term between executive constraints and
GDP per Capita I calculate the marginal effect of $\theta$, executive
constraints, on the predicted probability of civil conflict onset for
different levels of GDP per capita while holding constant the other
covariates. In this case, I take the mean of GDP per capita growth,
CINC, the proportion of discriminated population, population density,
mean elevation, and the number of politically relevant ethnic groups
rounded to the nearest integer. I also set the dummy indicators for
independence, major ongoing conflict, and neighbouring conflict to
zero following the majority of observations in the dataset. I set the
random intercepts for year and country to zero, which is the average
effect. For peace years I take the intercept corresponding to the
average number of peace years rounded to the nearest integer. The
levels of GDP per capita were select based on the 10\%, 50\%, and 90
\% quantiles of the standardized log transformed values. These levels
roughly correspond to the GDP per capita of Rwanda in 2007 at
$`r gdp_rwa`$ 2011 US dollars, the Philippines in 2008 at $`r gdp_phl`$
2011 US dollars, and Portugal in 2009 at $`r gdp_po`$ 2011 US
dollars. In the dataset there are 60 conflict episode onsets below the
first level, 152 between the first and second, 55 between the second
and third, and 4 above the third. The results are plotted in figure
\ref{fig:int}.

```{r fig.cap = "\\label{fig:int} Interaction between executive constraints and GDP per capita with 80 percent CIs"}
# Interaction plot
medians <- apply(theta, 2, quantile, probs = 0.5)
lower_bound <- min(medians) %>% round(digits = 1)
upper_bound <- max(medians) %>% round(digits = 1)

theta_seq <- seq(lower_bound, upper_bound, .01)
state <- quantile(X[, "rgdpepc"], probs = c(0.1, .5, 0.9))

@@

From the listed plots it is clear that for medium to high income
country-years the overall effect of executive constraints when
interacted with GDP per capita on the predicted probability of
conflict onset is negative --- that is to say a more constrained
executive is associated with a decreased risk of civil conflict onset.
For example, for country-years with a relatively high level of GDP per
capita, going from the worst to best in executive constraints
decreases the predicted probability from [1.5, 4.5] down to below
0.5. This effect diminishes, however, for lower values of GDP per
capita to the point that executive constraints are practically
indeterminate to the risk of conflict onset for country-years situated
around the 10\% quantile for GDP per capita due to the high degree of
posterior uncertainty. In the low GDP per capita plot, there is,
nonetheless, a slight increase in the median predicted probability,
which begs the question: for GDP per capita values lower than the 10\%
quantile is there a distinguishable reversal in the relationship
between executive constraints and civil conflict such that there is
actually a *positive* increase in the predicted probability for higher
levels of executive constraints?

```{r fig.cap="\\label{fig:gdp} Difference in predicted probabilities between the lowest and highest level of executive constraints for different values of GDP per capita"}
rgdpepc <- seq(min(X[, "rgdpepc"]), state[1], 0.01)
df <- mclapply(rgdpepc, function(y) {
    low <- boot::inv.logit(fn(lower_bound, y)) * 100
    high <- boot::inv.logit(fn(upper_bound, y)) * 100

    post_summarise(high - low, probs = c(0.1, 0.5, 0.9)) %>%
        mutate(y = y,
               prob = mean(high > low))
}, mc.cores = 4) %>% bind_rows

df %<>%
    rename(codelow = `10%`, median = `50%`, codehigh = `90%`) %>%
    mutate(interval = sprintf("[%.3f, %.3f]",
                              signif(codelow, 3),
                              signif(codehigh, 3)))

df$original <- exp(df$y * sd(final.df$rgdpepc) + mean(final.df$rgdpepc))

ggplot(df, aes(original, median)) +
    geom_line() +
    geom_ribbon(aes(ymin = codelow, ymax = codehigh), alpha = 0.2) +
    theme_tufte() +
    xlab("GDP per Capita")
```

To investigate this possibility, \ref{fig:gdp} plots the pointwise
difference between the predicted probability of civil conflict given
the highest level of executive constraints and the predicted
probability given the lowest level of executive constraints for
differing values of GDP per capita below the 10\% quantile and down to
the minimum observed value in the dataset while holding constant the
remaining covariates in the same fashion as the interaction plots in
figure \ref{fig:int}.[^theta_bounds] At the far end of the X-axis,
given a GDP per capita of approximately $1,109$ 2011 US dollars, there
is an 80 percent posterior probability that the difference lies within
the interval `r last(df$interval)`. At the opposite end, given a GDP
per capita of approximately $223$ 2011 US dollars, there is an 80
percent posterior probability that the difference is between
`r first(df$interval)`. The lower bounds of the credible intervals remain
fairly consistent across changing values of GDP per capita while the
upper bounds increases due to the distributions of predicted
probabilities being progressively skewed as GDP per capita
decreases. The high degree of uncertainty precludes the ability to
conclusively assert that there is a positive relationship between
civil conflict risk and executive constraints for low income
country-years. This is further bolstered by the fact that there are
relatively few observations falling within the extreme tail ends of
either the executive constraints posterior distribution or the
empirical distribution of GDP per capita --- in the case of the
latter, only `r signif(mean(X[, "rgdpepc"] < mean(df$y)), 3) * 100`
percent of the observations lay below the midpoint of the X-axis at
600 2011 US dollars.

[^theta_bounds]: Since the executive constraints parameter, $\theta$,
    is on the unbounded scale, the highest and lowest levels used in
    the calculations for \ref{fig:gdp} are drawn from the minimum and
    maximum of the median posterior estimates.

@@

@@

```{r psum}
p_median <- apply(p_hat, 2, median)

df <- bind_rows(summary(p_median[final.df$lepisode_onset == 1]),
                summary(p_median[final.df$lepisode_onset == 0])) %>%
    mutate_all(~signif(., 3))

@@

```{r confusion}
m <- table(p_median > .168, final.df$lepisode_onset)

@@

tpr <- signif(100 * m["Predicted: Onset", "Actual: Onset"] /
              sum(m[, "Actual: Onset"]), 3)
tnr <- signif(100 * m["Predicted: No Onset", "Actual: No Onset"] /
              sum(m[, "Actual: No Onset"]), 3)

fpr <- signif(100 * m["Predicted: Onset", "Actual: No Onset"] /
              sum(m[, "Actual: No Onset"]), 3)
fnr <- signif(100 * m["Predicted: No Onset", "Actual: Onset"] /
              sum(m[, "Actual: Onset"]), 3)

@@

To assess how well the presented model fits the observed data, I first
consider the predicted probabilities of conflict onset by plotting the
first 50 posterior draws for two groups, country-years that actually
experienced an onset of civil conflict and country-years that did
not. On average, the predicted probabilities for the actual onset
country-years are higher albeit still extremely low. This can be seen
in the summary statistics reported in table \ref{tab:psum} for the
posterior medians of the predicted probabilities. For onset
country-years the interquartile interval is only `r df[1, "1st Qu."]`
to `r df[1, "3rd Qu."]` with a max of `r df[1, "Max."]`. To better
compare the predicted probabilities to the actual data, table
\ref{tab:confusion} presents a confusion matrix with a 0.17 cutoff,
which approximately corresponds to the third quartile in the summary
table. `r tpr` percent of onsets and `r tnr` percent of non-onsets are
correctly predicted. The false positive rate is `r fpr` percent and
the false negative rate is `r fnr` percent.

```{r results = "hide"}
x <- evalmod(scores = p_median, labels = final.df$lepisode_onset)

@@

Figure \ref{fig:roc} plots a receiver operating characteristic (ROC)
curve depicting the tradeoff between the true positive rate
(*sensitivity*) and the false positive rate --- denoted by the formula
$1 - \mbox{specificity}$, where *specificity* refers to the true
negative rate --- for varying probability thresholds according to the
posterior medians of the predicted probabilities. For example, given a
threshold of approximately 0.085, the model will correctly predict
roughly 50 percent of onsets while misclassifying only 8 percent of
non-onsets. The total performance of the model can be summarized by
the area under the curve (AUC), which in this case is `r auc`.[^auc]
However, for the present analysis the results from the ROC curve are
misleading since there is a significant class imbalance in the
dataset. Onsets comprise only
`r signif(100 * mean(final.df$lepisode_onset == 1), 3)` percent of all
observations meaning that the false positive rate is less sensitive to
changing thresholds. Figure \ref{fig:roc} therefore also plots a
precision-recall (PRC) curve, which depicts the tradeoff between
*precision* --- the number of true positives divided by the sum of the
true positives and false positives --- and *recall* --- the number of
true positives divided by the sum of the true positives and false
negatives --- for varying thresholds. The advantage is that
*precision* is invariant to the number of non-onset years in the
dataset. From this plot it is clear that the precision of the model,
the ability to discriminate between onset country-years and falsely
predicted non-onset country-years, drops drastically as the *recall* of
the model increases. The *AUC-PR* for this plot is only
`r auc_pr`.[^auc_pr]

@@

Taken together, the evidence suggests a fairly poor fit to the
underlying data. The confusion matrix demonstrates that even with a
relatively low cutoff, the model is still only able to accurately
predict `r tpr` percent of conflict onsets. According to the ROC
curve, decreasing this threshold improves the true positive rate with
a relatively decent tradeoff in terms of the false positive rate, but
as stated, this is largely due to the imbalance in the outcome
variable. In terms of *precision*, the performance of the model
quickly degrades as the threshold decreases.

@@

```{stan, code = readLines("stan/sem.stan"), eval = F, output.var = "x", echo = T}
```
